---
title: "Class 18: Pertussis Mini Project"
author: "Yoonjin Lim (PID: A16850635)"
format: pdf
toc: true
---

## Background

Pertussis (a.k.a. whooping cough) is a common lung infection caused by the bacteria *B. Pertussis*. 

The CDC tracks cases of Pertussis in the U.S: 
https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html 

## Examining cases of Pertussis by year

We can use the **datapasta** package to scrape case numbers from the CDC website. 


```{r, include=FALSE}
cdc <- data.frame(
                          year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L,2024L),
  cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063, 35493)
)
```

> Q1. Make a plot of pertussis cases per year using ggplot 

```{r}
library(ggplot2)

cases <- ggplot(cdc) + 
  aes (year, cases) + 
  geom_point() + 
  geom_line()

cases
```

> Q2. Add some key time points in our history of interaction with Pertussis to our plot. These include wP roll-out (the first vaccine) in 1946 and the switch to aP in 1996. 

We can use `geom_vline()` for this. We can also use this function to add the key time point for COVID-19 in 2020. 

```{r}
cases + 
  geom_vline(xintercept=1946, col="blue") + 
  geom_vline(xintercept=1996, col="red") + 
  geom_vline(xintercept=2020, col="purple")
```

Yoonjin's Notes: Covid in 2020 => the first peak and after 2020.. another peak .. based on the observation. 
why cases number increases? 
- bacteria.. this bacteria doesn't seem to replicate crazy like virus.. 
- vaccine hesitancy
- socioeconomic factor 

However, what we observed was instead ... 
- Japan switched over their vaccine.. UK .. highly suggested something up related to this vaccine.. 

highly suggested that this new vaccine aP something about immune protection provided affected protection goes away faster ... based on the faster increased number of the cases. However, this should not be happened. 


> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Based on the graph above, it is clearly seen that the Pertussis cases have increased after the introduction of the aP vaccine from 1996. For example, it is observed that there is a high peak of Pertussis cases reported around 2012. Even though the cases has decreased after this peak, it is clear that this is a different trend observed when compared to the effect of wP vaccine. About this result, several different hypothesis has been suggested for the resurgence of Pertussis case: vaccination hesitancy, bacterial evolution escaped from vaccine immunity, socioeconomic factors, and weakened immunity in  puberty who got the vaccine as infants with the aP vaccine. However, it was instead highly suggested that something about immune protection provided from the newer aP vaccine impacts the immunity to go away faster based on the faster increase in the number of Pertussis cases, which is not supposed to happen.  

Mounting evidence suggests that the newer **aP** vaccine is less effective over the long term than than the older **wP** vaccine that it replaced. In other words, vaccine protection wanes more rapidly with ap than with wP...because the case number increase pretty rapidly after 1996. 

=> Immune responses for each vaccine is different


## Enter the CMI-PB project 

CMI-PB (computational Models of Immunity - Pertussis Boost) major goal is to investigate how the immune responds differently to with aP vs. wP vaccinated individuals and be able to predict this at an early stage.   

CMI-PB makes all their collected data freely available and they store it in a database composed different tables. Here we will access a few of these. 

We can use the **jsonlite** package to read this data. 

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", 
                     simplifyVector = TRUE)

head(subject)
```

> Q. How many subjects (i.e. enrolled people) are there in this dataset? 

```{r}
nrow(subject)
```

> Q4. How many "aP" and "wP" subjects are there? 

```{r}
subject$infancy_vac
```

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male/Female are in the dataset? 

```{r}
table(subject$biological_sex)
```

> Q6. How about gender and race numbers?

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Is this representative of the US population? 

NO. Definitely not. Because this is only a sub-population using few enrolled people, which doesn't represent all US population quantitatively. We need definitely more data for representing the US population. 

Notes: they use the year-of birth because age changes every single year. 

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)
today()
```
```{r}
today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


Let's read another database table from CMI-PB. 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector = TRUE)
ab_data <- read_json ("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",
                     simplifyVector=TRUE)
```

Wee peak at these. 

```{r}
head(specimen)
```

We want to "join" these tables to get all our information together. FOr this we will use **dplyr** package and the `inner_join()` function. 

> Q9_a. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

With the help of facet boxplot/historgram, I believe that these two groups wP and aP are significantly different in age. The aP group tends to be younger, whereas the wP group is older because the histogram of the aP group is more shifted to the left side, indicating younger people while the wP group's histogram is shifted toward the right side, representing older people. Based on the boxplot/histogram, I could clearly see the significant different in age distribution between these groups. I believe that this result is highly predicted because the aP vaccine replaced the wP vaccine in 1990s. 

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

Even the p-value of 2.372101e-23 illustrates that the probability of the age difference between the aP and the wP groups happened by just random chance is extremely low. This data supports highly statistically significant difference in age between the wP and aP groups. Since the vaccination type is likely related to the age, this suggests that the age and vaccination are not just independent.  

> Q9_b. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```

```{r}
dim(meta)
```

```{r}
head(ab_data)
```

One more "join" to get ab_data and meta all together. 

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_data, meta)
head(abdata)
```

```{r}
dim(abdata)
```

> Q11. How many Ab isotypes/antibodies are there in the dataset? 

```{r}
table(abdata$isotype)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

In abdata, the `abdata$dataset` column includes different data values from four years: 2020_dataset, 2021_dataset, 2022_dataset, and 2023_dataset. This different data values indicate specimens collected and analyzed in each of these years. Specifically, 31,520 specimens were collected from 2020. Similarly, 8,085 specimens, 7,301 specimens, and 15,050 specimens were collected and analyzed from each 2021, 2022, and 2023. Interestingly, the 2020_dataset has the largest number of rows, whereas the other year_dataset such as the most recent dataset 2023_dataset has fewer number of rows than the year 2020. However, the 2023_dataset has higher number of rows than 2021 and 2022. The number of specimens dropped in 2021 and 2022 mainly because of the COVID-19 pandemic disrupting regular lab and clinic operations. I believe that 2020_dataset still have the higher number because of collection of data from early-phase of the normal year. In 2023, the healthcare and clinical operation was stabilized but it would be able to collect the specimen partially, resulting in a partial increase in dataset size. 

> Q. How many different antigens are measured in the dataset? 

```{r}
table(abdata$antigen)
```

> Q. Make a boxplot of antigen levels across the whole dataset (MFI vs. antigen)? 

```{r}
ggplot(abdata) + 
  aes (MFI, antigen) + 
  geom_boxplot() 
```

> Q. Are there obvious differences between aP and wP values? 

```{r}
ggplot(abdata) + 
  aes (MFI, antigen, col=infancy_vac) + 
  geom_boxplot()
```

## Focus on IgG levels 

IgG is the most abundance antibody in blood. With four sub-classes (IgG1 to IgG4) crucial for long-term immunity and responding to bacterial & viral infections. 

```{r}
igg <- abdata |> 
  filter(isotype == "IgG")

head(igg)
```

Same boxplot of antigens as before

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

```{r}
ggplot(igg) + 
  aes (MFI_normalised, antigen, col = infancy_vac) + 
  geom_boxplot() + 
  facet_wrap(~visit)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

Based on Q13 with the figure above, this boxplot basically illustrates the IgG antibody levels (MFI_normalised) against several different Pertussis-related antigens, grouped by visit number (1 to 8). From this figure, I could observe that there are several different antigens that shows significant increase in the level of IgG antibody titers, indicating strength of immune response: PT, PRN, and FIM2/3. These antigens illustrates the significant increase of IgG levels around visit 5-7, highly indicating a strong immune response after vaccination or against the exposure. In particular, the antigen FIM2/3 shows the most dramatic increase of IgG levels by larger boxplot. FHA also illustrates the increase of IgG levels around the same time, but not as much as of PT, PRN, and FIM2/3. On the other hand, there are other antigens such as OVA, TT, and DT that doesn't illustrate the dramatic change among the visits because I believe that they are not likely to be direct targeted by Pertussis vaccine, suggesting how the immune response are specifically reacted to particular vaccine components.



Focus in further in just one of these antigens - let's pick **PT** (Pertussis Toxin, one of the main toxins of the bacteria) in the **2021_dataset** again for **IgG** antibody isotypes. 

```{r}
table(igg$dataset)
```

```{r}
pt_igg <- abdata |> 
  filter (isotype == "IgG",
          antigen == "PT", 
          dataset == "2021_dataset")
```

```{r}
dim(pt_igg)
```

```{r}
ggplot(pt_igg) + 
  aes(actual_day_relative_to_boost, 
      MFI_normalised, 
      col=infancy_vac, 
      group = subject_id) + 
  geom_point() + 
  geom_line () + 
  theme_bw() + 
  geom_vline(xintercept=0) + 
  geom_vline(xintercept=14)
```

Notes: join the all the point/line together = grouping / peak levels. What does this represent? 
the other vaccine respond differently.. immune response... 



